---
title: "EDA"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

# Final Project 01: EDA

[Youth Risk Behavior Surveillance System (YRBSS), 2023](https://www.cdc.gov/yrbs/data/index.html)

## Exploratory data analysis

```{r, message = FALSE}
library(tidyverse)
library(Hmisc)
library(naniar)

mdb.get('XXH2023_YRBS_Data.mdb', tables = TRUE)

# full data set will all item responses
data <- mdb.get('XXH2023_YRBS_Data.mdb', tables = "XXHq")

# already pre-processed to make each question dichotomous
data_qn<- mdb.get('XXH2023_YRBS_Data.mdb', tables = "XXHqn")

# see data dictionary (pdf included in project folder) for individual item responses and question wording. 

vars <- data |> select( 
  q1,  # age
  q2,  # sex
  q19, # forced sexual intercourse (lifetime)
  q20, # sexual violence (12 months)
  q21, # sexual dating violence (12 months)
  q22, # physical dating violence (12 months)
  q33, # current cigarette use (30 days)
  q36, # current electronic vapor use (30 days)
  q42, # current alcohol use (30 days)
  q48, # current marijuana use (30 days)
  q57, # age first sexual intercourse
  q58, # number sexual partners (lifetime)
  q59, # current sexual activity/number sexual partners (3 months)
  q60, # alcohol/drug use during sex (last time)
  q61, # condom use (last time)
  q62, # birth control method use (last time)
  q64, # sexual orientation
  q80, # social media use
  q81, # HIV testing (lifetime)
  q82, # STD testing (12 months)
  q84  # current mental health (30 days)
  ) 
```

## What's in my data?

```{r}
str(data)
```

Each of the questions asked in the National High School Youth Risk Behavior Survey (YRBS) has its own column in the dataset. Each row represents a participant in the survey, so their responses associated to each of questions is a single observation. The dataset mostly consists of integers that correlate to each of the possible multiple choice answers. So, if for question 1 (q1), when asked how old they were, the observation is 1. That would mean that person responded with A), they were 12 years or younger.

## Data Variation

```{r, warning=FALSE}
# demographic data variation
ggplot(data = data) +
  geom_histogram(mapping = aes(x= q84), bins = 5, color = "navy", fill = "lightblue")+
  labs(x = "Mental Health",
       y = "Count",
       title = "Mental Health Histogram") 
```

Frequency of poor mental health (1 - Never, 2 - Rarely, 3 - Sometimes, 4 - Most of the time, 5 - Always) looks normally distributed, perhaps with a slight right skew. 

```{r, warning=FALSE}
# first drink
ggplot(data = data) +
  geom_histogram(mapping = aes(x= q42), bins = 7, color = "red", fill = "orange")+
  labs(x = "Frequency",
       y = "Count",
       title = "Distribution of Current Alcohol Use") 
```

However, for distribution of current alcohol use there was dramatic right skew as the vast majority responding to the survey haven't consumed alcohol. This is unsurprising given that alcohol use is a lower frequency behavior, let along significant/frequent use. 

## Missing Data

```{r}
# missing data for complete data
summary(vars)
vis_miss(vars, warn_large_data = FALSE)
```

There is a noticeable amount of missing data (12.3%), which is clearly evident in a missing data visualization. Importantly, missing data does not seem to be at random, with specific observations having large portions of missing data. Implications of missing data and steps for corrective action are outlined in the project description.  

## Covariation
```{r}
# take numeric data out 
numeric_data <- vars[sapply(vars, is.numeric)]
corr_matrix <- cor(numeric_data, use = "complete.obs", method = "pearson")
print(corr_matrix)
```

```{r}
library(reshape2)
melted_corr <- melt(corr_matrix)

# find strongly correlated pairs
strong_corrs <- subset(melted_corr, Var1 != Var2 & (value > 0.7 | value < -0.7))

# get rid of duplicate pairs
strong_corrs_unique <- strong_corrs[!duplicated(t(apply(strong_corrs[ , 1:2], 1, sort))), ]

# View the results
print(strong_corrs_unique)
nrow(strong_corrs_unique)
```

There are 12 pairs of correlated variables.

# Project description

This project is focused on methodology, rather than attempting to answer a specific question about the data itself. Instead, the learning outcomes for this project are centered around gaining intuition for different unsupervised classification models and their appropriate uses. The core component of this project is building multiple types of models and comparing the strengths and appropriateness of their classification models. The models included will be primarily unsupervised (K-means clustering, hierarchical clustering), but will also include a supervised model (decision tree). For this project to successful, pre-processing of data and careful selection of predictor variables is essential. Based on the initial exploratory data analysis conducted here, a potential problem is the amount of missing data in the data set. Missing data analysis (e.g., MCAR, MAR, MNAR) and imputation are important components of model selection, and does impact both the types of models fit as well as the kinds of conclusions that can be drawn. However, evaluating and compensating for missing data like the type seen in this data set is beyond the scope of this project both due to time constraints. For this project, the data set will be significantly reduced using list wise deletion with the understanding that this is a dramatic oversimplification of the data and results would likely vary if other methods were used to deal ith missing data.
