---
title: "EDA"
format: pdf
editor: visual
editor_options: 
  chunk_output_type: console
---

# Final Project 01: EDA

[Youth Risk Behavior Surveillance System (YRBSS), 2023](https://www.cdc.gov/yrbs/data/index.html) https://www.cdc.gov/yrbs/data/index.html

## Exploratory data analysis
```{r}
library(tidyverse)
library(Hmisc)
library(naniar)

mdb.get('XXH2023_YRBS_Data.mdb', tables = TRUE)

# full data set will all item responses
data <- mdb.get('XXH2023_YRBS_Data.mdb', tables = "XXHq")

# already pre-processed to make each question dichotomous
data_qn<- mdb.get('XXH2023_YRBS_Data.mdb', tables = "XXHqn")
```

## Whats in my data

## Data variation
Visualize distributions
 
## Missing Data
```{r}
# missing data for complete data
summary(data)
vis_miss(data, warn_large_data = FALSE)
```
There is a large amount of missing data, which is clearly evident in a missing data visualization. Around 50% of observations are missing data for most of the questions, with certain questions having no data whatsoever. This is likely due to the nature in which the data was collected. As this is national school based survey, 

```{r}
# missing data for QN data
summary(data_qn)
vis_miss(data_qn, warn_large_data = FALSE)
```
Compared to the full data set, there is more missing data for the calculated dichotomous variables. 


## Covariation

# Project description
  This project is focused on methodology, rather than attempting to answer a specific question about the data itself. Instead, the learning outcomes for this project are centered around gaining intuition for different unsupervised classification models and their appropriate uses.
  The core component of this project is building multiple types of models and comparing the strengths and appropriateness of their classification models. The models included will be primarily unsupervised (K-means clustering, hierarchical clustering), but will also include a supervised model (decision tree). For this project to successful, pre-processing of data and careful selection of predictor variables is essential. 
  Based on the initial exploratory data analysis conducted here, a potential problem is the amount of missing data in the data set. Missing data analysis (e.g., MCAR, MAR, MNAR) and imputation are important components of model selection, and does impact both the types of models fit as well as the kinds of conclusions that can be drawn. However, evaluating and compensating for missing data like the type seen in this data set is beyond the scope of this project both due to time constraints. For this project, the data set will be significantly reduced using list wise deletion with the understanding that this is a dramatic oversimplification of the data and results would likely vary if other methods were used to deal ith missing data. 
